---
title: "Sperm Long RNA Analysis"
output: html_document
author: "Bohan Zhang"
date: "2026-06-03"
---

```{r setup}
library(edgeR)
library(SummarizedExperiment)
library(clusterProfiler)
library(org.Mm.eg.db)
library(ggplot2)
library(limma)
library(ComplexHeatmap)
library(sechm)
library(decoupleR)
library(tidyr)
library(rtracklayer)
library(epiwraps)
library(variancePartition)
library(BiocParallel)
library(openxlsx)
library(ggpubr)
library(emmeans)
library(GenomicFeatures)
library(txdbmaker)
library(AnnotationHub)
library(GenomeInfoDb)
library(memes)
library(ggrepel)
library(cowplot)
library(patchwork)
library(magick)
library(cluster)
library(ggsignif)
library(ggplotify)
ah <- AnnotationHub()
setwd("/mnt/gapp/work/bohan/thesis/IFS_sperm_longRNA")
```

# 1. DEA with 2 SVs

```{r}
se <- readRDS("2026_iryna_EXP4_bohan_IFS_PROTAC_longRNA.SE.SV2.rds")
metadata(se)$anno_colors <- list(
  stress=c(ctrl="steelblue", ifs ="firebrick3"), 
  protac = c(PROTAC = "orange2", DMSO = "navy"),
  group = c("ctrl_DMSO" = "seashell2", "ifs_DMSO" =  "#ffa500ff" , "ctrl_PROTAC" = "seashell4" , "ifs_PROTAC" = "tomato3"))


dge_for_filter <- DGEList(assay(se, "counts"))
dge_for_filter <- calcNormFactors(dge_for_filter)
design_for_filter <- model.matrix(~ 0 + group, data = as.data.frame(colData(se)))
# Filter using the temporary design matrix
keep <- filterByExpr(dge_for_filter, min.count = 20, design = design_for_filter)
se  <- se[keep, ]
# add donor info
colData(se)$donor <- sub("^(C[0-9]+|I[0-9]+)[DK]$", "\\1", colData(se)$sampleid)
colData(se)$donor <- factor(colData(se)$donor)

# voom
dge <- DGEList(assay(se, "counts"))
dge <- calcNormFactors(dge)
design <- model.matrix(~ 0 + group + SV1 + SV2, data = as.data.frame(colData(se)))
assay(se, "logcpm") <- log1p(edgeR::cpm(dge)[rownames(se),])

# в”Ђв”Ђ 2. First voom + first duplicateCorrelation pass в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
vobj_for_dupcor <- voom(dge, design, plot = FALSE)
dupcor <- duplicateCorrelation(vobj_for_dupcor, design, block = colData(se)$donor)

# в”Ђв”Ђ 3. Second voom pass, now informed by the donor correlation в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
vobj <- voom(dge, design, plot = FALSE,
             block       = colData(se)$donor,
             correlation = dupcor$consensus)

# в”Ђв”Ђ 4. Second duplicateCorrelation pass on the better voom weights в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
dupcor <- duplicateCorrelation(vobj, design, block = colData(se)$donor)

# в”Ђв”Ђ 5. Fit the linear model в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
fit <- lmFit(vobj, design, block=colData(se)$donor, correlation = dupcor$consensus)

contrast <- makeContrasts(
  total_stress_vs_ctrl  = (groupifs_DMSO + groupifs_PROTAC)/2 - (groupctrl_DMSO + groupctrl_PROTAC)/2,
  DMSO_stress_vs_ctrl   = groupifs_DMSO - groupctrl_DMSO,
  PROTAC_stress_vs_ctrl = groupifs_PROTAC - groupctrl_PROTAC,
  total_PROTAC_vs_DMSO  = (groupifs_PROTAC + groupctrl_PROTAC)/2 - (groupifs_DMSO + groupctrl_DMSO)/2,
  ctrl_PROTAC_vs_DMSO   = groupctrl_PROTAC - groupctrl_DMSO,
  stress_PROTAC_vs_DMSO = groupifs_PROTAC - groupifs_DMSO,
  interaction           = (groupifs_PROTAC - groupctrl_PROTAC) - (groupifs_DMSO - groupctrl_DMSO),
  levels = design
)

fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)  # Empirical Bayes smoothing of standard errors

# extract each contrast table 
rowData(se)$DEA.voom.total_stress_vs_ctrl <- topTable(fit2, coef = "total_stress_vs_ctrl", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.DMSO_stress_vs_ctrl <- topTable(fit2, coef = "DMSO_stress_vs_ctrl", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.PROTAC_stress_vs_ctrl <- topTable(fit2, coef = "PROTAC_stress_vs_ctrl", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.total_PROTAC_vs_DMSO <- topTable(fit2, coef = "total_PROTAC_vs_DMSO", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.ctrl_PROTAC_vs_DMSO <- topTable(fit2, coef = "ctrl_PROTAC_vs_DMSO", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.stress_PROTAC_vs_DMSO <- topTable(fit2, coef = "stress_PROTAC_vs_DMSO", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.interaction <- topTable(fit2, coef = "interaction", n = Inf)[rownames(se),]

```

## Fig 12A

Heatmap of DEG result of SVA corrected expression

```{r}
results_total_stress_vs_ctrl <- rowData(se)$DEA.voom.total_stress_vs_ctrl
sig_total_stress_vs_ctrl <- results_total_stress_vs_ctrl[results_total_stress_vs_ctrl$adj.P.Val<0.05,]
table(sign(sig_total_stress_vs_ctrl$logFC))
sig_total_stress_vs_ctrl <- rownames(sig_total_stress_vs_ctrl)
table(rownames(se)==rownames(vobj$E)) # they have the same order
assay(se, "voom.val") <- vobj$E
```

corrected value

```{r} 
# pdf("heatmap_sperm_longRNA_voomDEG_total_stress_vs_ctrl.pdf", width = 7, height = 7)
fig12a <- sechm(se, 
      sig_total_stress_vs_ctrl, 
      assayName = "corrected",
      top_annotation = c("stress", "protac"),
      gaps_at = c("stress"),
      cluster_rows = TRUE,
      do.scale = TRUE,
      column_title = "Heatmap of SVA corrected value", 
      column_title_gp = gpar(fontsize = 18, fontface = "bold"),
      show_annotation_legend = FALSE,
      show_heatmap_legend = FALSE,
      use_raster = TRUE)
# draw(fig12a)
# dev.off()
fig12a <- grid::grid.grabExpr(draw(fig12a))
```

## Fig 12B

volcano plot of DEA result of SVA corrected expression

```{r}
results_total_stress_vs_ctrl <- results_total_stress_vs_ctrl[order(results_total_stress_vs_ctrl$adj.P.Val),]
results_total_stress_vs_ctrl$gene <- rownames(results_total_stress_vs_ctrl)

results_total_stress_vs_ctrl$significance <- "non"
results_total_stress_vs_ctrl$significance[results_total_stress_vs_ctrl$adj.P.Val < 0.05 & results_total_stress_vs_ctrl$logFC > 0] <- "up"
results_total_stress_vs_ctrl$significance[results_total_stress_vs_ctrl$adj.P.Val < 0.05 & results_total_stress_vs_ctrl$logFC < 0] <- "down"

top5updown <- rbind(
  head(results_total_stress_vs_ctrl[results_total_stress_vs_ctrl$significance == "up", ], 5),
  head(results_total_stress_vs_ctrl[results_total_stress_vs_ctrl$significance == "down", ], 5)
)

fig12b <- ggplot(results_total_stress_vs_ctrl, aes(x = logFC, y = -log10(adj.P.Val), label = gene, color = significance)) +
  geom_point() +
  geom_text_repel(
    data = top5updown,
    size = 5,
    max.overlaps  = 20,
    box.padding   = 0.3,
    color = "black"
    ) +
  scale_color_manual(values = c("up" = "red", "non" = "gray", "down" = "blue")) +
  labs(x = "logFC Stress vs. Ctrl", y = "-log10(Padj)", title = "Volcano plot of sperm long RNA DEG result") +
  theme_bw()+
  theme(
    axis.title = element_text(size = 18),
    axis.text  = element_text(size = 16),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.text  = element_text(size = 16),
    legend.title = element_text(size = 16)
  )
fig12b

# ggsave("./plot_for_thesis/volcano_sperm_long_voomDEG_total_stress_vs_ctrl.pdf", fig12b, width = 7, height = 7)
```

## Fig 12c

logcpm (uncorrected value)

```{r}
# for thesis
# pdf("heatmap_sperm_longRNA_logcpm_voomDEG_total_stress_vs_ctrl.pdf", width = 7, height = 7)
fig12c <- sechm(se, 
      sig_total_stress_vs_ctrl, 
      assayName = "logcpm",
      top_annotation = c("stress", "protac"),
      gaps_at = c("stress"),
      cluster_rows = TRUE,
      do.scale = TRUE,
      use_raster = TRUE,
      column_title = "Heatmap of logcpm (uncorrected value)",
      column_title_gp = gpar(fontsize = 18, fontface = "bold"))
# draw(fig12c)
# dev.off()

fig12c <- grid::grid.grabExpr(draw(fig12c))
```


# 2. DEA without sva correction

```{r}
se <- readRDS("2026_iryna_EXP4_bohan_IFS_PROTAC_longRNA.SE.rds")

metadata(se)$anno_colors <- list(
  stress=c(ctrl="steelblue", ifs ="firebrick3"), 
  protac = c(PROTAC = "orange2", DMSO = "navy"),
  group = c("ctrl_DMSO" = "seashell2", "ifs_DMSO" =  "#ffa500ff" , "ctrl_PROTAC" = "seashell4" , "ifs_PROTAC" = "tomato3"))

se$group <- factor(paste0(se$stress, "_", se$protac))

dge_for_filter <- DGEList(assay(se, "counts"))
dge_for_filter <- calcNormFactors(dge_for_filter)
design_for_filter <- model.matrix(~ 0 + group, data = as.data.frame(colData(se)))
# Filter using the temporary design matrix
keep <- filterByExpr(dge_for_filter, min.count = 20, design = design_for_filter)
se  <- se[keep, ]
# add donor info
colData(se)$donor <- sub("^(C[0-9]+|I[0-9]+)[DK]$", "\\1", colData(se)$sampleid)
colData(se)$donor <- factor(colData(se)$donor)

# voom
dge <- DGEList(assay(se, "counts"))
dge <- calcNormFactors(dge)
design <- model.matrix(~ 0 + group, data = as.data.frame(colData(se)))

# also calculate logcpm
assay(se, "logcpm") <- log1p(edgeR::cpm(dge)[rownames(se),])

# в”Ђв”Ђ 2. First voom + first duplicateCorrelation pass в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
vobj_for_dupcor <- voom(dge, design, plot = FALSE)
dupcor <- duplicateCorrelation(vobj_for_dupcor, design, block = colData(se)$donor)

# в”Ђв”Ђ 3. Second voom pass, now informed by the donor correlation в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
vobj <- voom(dge, design, plot = FALSE,
             block       = colData(se)$donor,
             correlation = dupcor$consensus)

# в”Ђв”Ђ 4. Second duplicateCorrelation pass on the better voom weights в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
dupcor <- duplicateCorrelation(vobj, design, block = colData(se)$donor)

# в”Ђв”Ђ 5. Fit the linear model в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
fit <- lmFit(vobj, design, block=colData(se)$donor, correlation = dupcor$consensus)

contrast <- makeContrasts(
  total_stress_vs_ctrl  = (groupifs_DMSO + groupifs_PROTAC)/2 - (groupctrl_DMSO + groupctrl_PROTAC)/2,
  DMSO_stress_vs_ctrl   = groupifs_DMSO - groupctrl_DMSO,
  PROTAC_stress_vs_ctrl = groupifs_PROTAC - groupctrl_PROTAC,
  total_PROTAC_vs_DMSO  = (groupifs_PROTAC + groupctrl_PROTAC)/2 - (groupifs_DMSO + groupctrl_DMSO)/2,
  ctrl_PROTAC_vs_DMSO   = groupctrl_PROTAC - groupctrl_DMSO,
  stress_PROTAC_vs_DMSO = groupifs_PROTAC - groupifs_DMSO,
  interaction           = (groupifs_PROTAC - groupctrl_PROTAC) - (groupifs_DMSO - groupctrl_DMSO),
  levels = design
)

fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)  # Empirical Bayes smoothing of standard errors

# extract each contrast table 
rowData(se)$DEA.voom.total_stress_vs_ctrl <- topTable(fit2, coef = "total_stress_vs_ctrl", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.DMSO_stress_vs_ctrl <- topTable(fit2, coef = "DMSO_stress_vs_ctrl", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.PROTAC_stress_vs_ctrl <- topTable(fit2, coef = "PROTAC_stress_vs_ctrl", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.total_PROTAC_vs_DMSO <- topTable(fit2, coef = "total_PROTAC_vs_DMSO", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.ctrl_PROTAC_vs_DMSO <- topTable(fit2, coef = "ctrl_PROTAC_vs_DMSO", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.stress_PROTAC_vs_DMSO <- topTable(fit2, coef = "stress_PROTAC_vs_DMSO", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.interaction <- topTable(fit2, coef = "interaction", n = Inf)[rownames(se),]

```

## Fig 13A

PROTAC effect on all GR binding long RNA in sperm 

```{r}
mmu_sig_rip_by_OmicsBridge <- read.xlsx("mmu_rip_gene_name.xlsx", colNames = FALSE)
# calculate the logFC manually to avoid anything to do with SVA
se <- log2FC(se, fromAssay = "logcpm", controls = se$protac == "DMSO", by = se$stress, isLog = TRUE)

allGR_se <- se[rownames(se)%in% mmu_sig_rip_by_OmicsBridge$X1, ]
# because here we calculate logFC(protac/dmso), and within the stress group. So here we can just use the avg logFC of all the protac samples as the final logFC for this gene
df_allGR_logFC <- rowMeans(assay(allGR_se, "log2FC")[,allGR_se$protac=="PROTAC"])
n_down <- sum(df_allGR_logFC<0)
N <- length(df_allGR_logFC)
binom.test(n_down, N, p = 0.5, alternative = "two.sided")
# to plot the heatmap, use the normalization to donor
allGR_se_for_heatmap <- allGR_se
allGR_se_for_heatmap <- log2FC(allGR_se_for_heatmap, fromAssay = "logcpm", controls = allGR_se_for_heatmap$protac == "DMSO", by = allGR_se_for_heatmap$donor, isLog = TRUE)

rowData(allGR_se_for_heatmap)$direction <- ifelse(df_allGR_logFC>0,yes = "up",no = "down")
```

```{r}
# pdf("./plot_for_thesis/longRNA_PROTAC_effect.pdf", width=6, height=8)
fig13a <- sechm(
  allGR_se_for_heatmap[,allGR_se_for_heatmap$protac == "PROTAC"],
  names(df_allGR_logFC),
  assayName = "scaledLFC",
  cluster_rows = F,
  top_annotation = c("protac"),
  # hmcols =colorRampPalette(c("blue", "black", "orange"))(101),
  gaps_at = c("protac"),
  gaps_row = "direction",
  column_title = "PROTAC effect on GR binding long RNA", 
  use_raster = TRUE,
  column_title_gp = gpar(fontsize = 18, fontface = "bold"),
  show_annotation_legend = FALSE
  ) 
# draw(fig13a)
# dev.off()
fig13a <- grid::grid.grabExpr(draw(fig13a))
```

## Fig 13B

PROTAC effect on stress affected GR binding long RNA (punadj<0.01)

```{r}
semisig <- rowData(se)$DEA.voom.total_stress_vs_ctrl[rowData(se)$DEA.voom.total_stress_vs_ctrl$P.Value<0.01, ]
GR_semisig <- intersect(rownames(semisig), mmu_sig_rip_by_OmicsBridge$X1) # 67

se_semisig <- se[rownames(semisig),]  # 952 genes
logFC_general <- rowMeans(assay(se_semisig, "log2FC")[,se_semisig$protac=="PROTAC"])
p_general <- sum(logFC_general<0)/length(logFC_general)

se_GR_semisig <- se[GR_semisig, ]  # 67
df_se_GR_semisig_logFC <- rowMeans(assay(se_GR_semisig, "log2FC")[,se_GR_semisig$protac=="PROTAC"])
n_down <- sum(df_se_GR_semisig_logFC<0)  # 60
N <- length(df_se_GR_semisig_logFC)  # 67

binom.test(n_down, N, p = p_general, alternative = "two.sided")

```

```{r}
# pdf("./plot_for_thesis/longRNA_with_p0.01_PROTAC_effect.pdf", width=6, height=8)
fig13b <- sechm(
  allGR_se_for_heatmap[,allGR_se_for_heatmap$protac == "PROTAC"],
  GR_semisig,
  show_rownames = T,
  assayName = "scaledLFC",
  cluster_rows = F,
  top_annotation = c("protac"),
  sortRowsOn = colnames(allGR_se_for_heatmap[,allGR_se_for_heatmap$protac == "PROTAC"]),
  gaps_row = "direction",
  # hmcols =colorRampPalette(c("blue", "black", "orange"))(101),
  column_title = "PROTAC effect on stress affected\nGR binding long RNA (p < 0.01)", 
  use_raster = TRUE,
  show_row_dend = FALSE,
  row_names_gp = grid::gpar(fontsize = 12),
  column_title_gp = gpar(fontsize = 18, fontface = "bold")
  ) 
# draw(fig13b)
# dev.off()
fig13b <- grid::grid.grabExpr(draw(fig13b))
```

# 3. check if the GR binders have GRE in their promoters

```{r}
semisig <- rowData(se)$DEA.voom.total_stress_vs_ctrl[rowData(se)$DEA.voom.total_stress_vs_ctrl$P.Value<0.01, ]
mmu_sig_rip_by_OmicsBridge <- read.xlsx("mmu_rip_gene_name.xlsx", colNames = FALSE)
mmu_sig_rip_by_OmicsBridge <- mmu_sig_rip_by_OmicsBridge[!duplicated(mmu_sig_rip_by_OmicsBridge$X1),]
mmu_sig_rip_by_OmicsBridge <- mmu_sig_rip_by_OmicsBridge[mmu_sig_rip_by_OmicsBridge %in% rownames(se)]

GRbindingRNA_sig <- intersect(mmu_sig_rip_by_OmicsBridge, rownames(semisig))

# # extract the promoters
# genome <- ah[["AH89457"]]        # EnsDb for Ensembl, GRCm39
# txs <- transcripts(genome)        # Extract all transcripts
# txs <- keepStandardChromosomes(txs, pruning.mode = "coarse")  # Remove non-standard contigs
# txs <- dropSeqlevels(txs, "MT", pruning.mode = "coarse")      # Remove mitochondrial genes
# seqnames(txs)

# # Get promoters
# proms <- promoters(txs, upstream = 1000, downstream = 100)

# seqlevelsStyle(proms)
# seqlevelsStyle(proms) <- "Ensembl"

# # Get GRE motif
# genome <- ah[["AH90965"]]. # TwoBitFile object

# GRE_motif <- MotifDb::MotifDb %>% 
#   MotifDb::query("GRE") %>% 
#   universalmotif::convert_motifs() %>% 
#   .[[1]]

# prom_seqs <- get_sequence(proms, genome)
# fimo <- runFimo(prom_seqs, GRE_motif, meme_path = "/common/meme/bin")

# GRE_proms <- proms[overlapsAny(proms, fimo)]

# all above can be replaced by 
GRE_proms <- readRDS("/mnt/gapp/work/miriam/EXP14_accessibility_sperm_embryo/rds_saves/GRE_promoters.rds")

# add symbol ID
library(org.Mm.eg.db)
symbols <- mapIds(
  org.Mm.eg.db,
  keys    = GRE_proms$gene_id,
  column  = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)
GRE_proms$symbol <- symbols[GRE_proms$gene_id]

GRbindingRNA_all_GRE_promoter <- intersect(GRE_proms$symbol, mmu_sig_rip_by_OmicsBridge)
# 671 genes in total 

GRbindingRNA_sig_GRE_promoter <- intersect(GRE_proms$symbol, GRbindingRNA_sig)
# [1] "Retreg2" "Atg9a"   "Asb1"    "Mbd3"   
# [5] "Git1"    "Tex14"   "Rnf44"   "Sorbs3" 
# [9] "Foxred2" "Elfn2"   "Espl1"   "Syngap1"
# [13] "Ddx11"   "Tpgs2"   "Fam229a" "Wdtc1"  
# [17] "Spata21" "Stag3"   "Cpa5"    "Bcl2l14"
# [21] "Egln2"   "Yif1b"   "Gpi1"    "Ptdss2" 
# [25] "Got2"    "Pde4a"   "Gramd2a" "Pcbp4"  

GRbindingRNA_sig_GRE_promoter_DEA <- semisig[GRbindingRNA_sig_GRE_promoter, ]
# write.xlsx(GRbindingRNA_sig_GRE_promoter_DEA, "GRbindingRNA_sig_GRE_promoter_DEA.xlsx")
```

# 4. check if the DEGs express only in brain and testis, but not elsewhere

```{r}
mgi <- readRDS("/mnt/gapp/work/bohan/thesis/IFS_sperm_longRNA/MGI_gxdrnaseq_TPMmatrix.rds")
brainOrTestis <- "brain|testis|hippo|cereb|cephalon|nucleus|bulb|thalamus|cortex|gyrus|corpus"
brainOrTestisMax <- rowMaxs(mgi[,grep(brainOrTestis, colnames(mgi))])
otherMax <- rowMaxs(mgi[,grep(paste0(brainOrTestis,"|spinal"), colnames(mgi), invert=TRUE)])
lfc <- log2((0.1+brainOrTestisMax)/(0.1+otherMax))
brainOrTestisSpecific <- otherMax<1 & brainOrTestisMax>1 & lfc>=1
intersect(rownames(semisig), names(brainOrTestisSpecific)[brainOrTestisSpecific])
```


# 5. circRNA - no SVA

```{r}
circRNA_bsj <- read.csv("/mnt/gapp/work/bohan/thesis/IFS_sperm_longRNA/prep_CIRIquant_output/circRNA_bsj.csv")
circRNA_info <- read.csv("/mnt/gapp/work/bohan/thesis/IFS_sperm_longRNA/prep_CIRIquant_output/circRNA_info.csv")
rownames(circRNA_bsj) <- circRNA_bsj$circ_id
rownames(circRNA_info) <- circRNA_info$circ_id
# circRNA_bsj$gene_name <- circRNA_info[rownames(circRNA_bsj),]$gene_name

circ_counts <- circRNA_bsj[, -1]
se <- SummarizedExperiment(
  assays = list(counts = circ_counts),
  colData = data.frame(sampleid = colnames(circ_counts), 
                 stress = ifelse(substr(colnames(circ_counts), 1, 1)=="C", "ctrl", "ifs"),
                 protac = ifelse(substr(colnames(circ_counts), nchar(colnames(circ_counts)), nchar(colnames(circ_counts)))=="D", "DMSO", "PROTAC"))
)
se$group <- factor(paste0(se$stress, "_", se$protac))

# add the host gene info to se
rowData(se)$host_gene <- circRNA_info[rownames(se),]$gene_name
rowData(se)$circ_type <- circRNA_info[rownames(se),]$circ_type
rowData(se)$gene_type <- circRNA_info[rownames(se),]$gene_type
rowData(se)$strand <- circRNA_info[rownames(se),]$strand


dge_for_filter <- DGEList(assay(se, "counts"))
dge_for_filter <- calcNormFactors(dge_for_filter)
design_for_filter <- model.matrix(~ 0 + group, data = as.data.frame(colData(se)))
keep <- filterByExpr(dge_for_filter, min.count = 5, design = design_for_filter)  # only use min.count=5 because circRNA has low reads.

se <- se[keep, ] # 4858 circRNAs kept

# voom
dge <- DGEList(assay(se, "counts"))
dge <- calcNormFactors(dge)
design<- model.matrix(~ 0 + group, data = as.data.frame(colData(se)))
assay(se, "logcpm") <- log1p(edgeR::cpm(dge)[rownames(se),])

# add donor info
colData(se)$donor <- sub("^(C[0-9]+|I[0-9]+)[DK]$", "\\1", colData(se)$sampleid)
colData(se)$donor <- factor(colData(se)$donor)

# в”Ђв”Ђ 2. First voom + first duplicateCorrelation pass в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
vobj_for_dupcor <- voom(dge, design, plot = FALSE)
dupcor <- duplicateCorrelation(vobj_for_dupcor, design, block = colData(se)$donor)

# в”Ђв”Ђ 3. Second voom pass, now informed by the donor correlation в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
vobj <- voom(dge, design, plot = FALSE,
             block       = colData(se)$donor,
             correlation = dupcor$consensus)

# в”Ђв”Ђ 4. Second duplicateCorrelation pass on the better voom weights в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
dupcor <- duplicateCorrelation(vobj, design, block = colData(se)$donor)

# в”Ђв”Ђ 5. Fit the linear model в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
fit <- lmFit(vobj, design, block=colData(se)$donor, correlation = dupcor$consensus)

contrast <- makeContrasts(
  total_stress_vs_ctrl  = (groupifs_DMSO + groupifs_PROTAC)/2 - (groupctrl_DMSO + groupctrl_PROTAC)/2,
  DMSO_stress_vs_ctrl   = groupifs_DMSO - groupctrl_DMSO,
  PROTAC_stress_vs_ctrl = groupifs_PROTAC - groupctrl_PROTAC,
  total_PROTAC_vs_DMSO  = (groupifs_PROTAC + groupctrl_PROTAC)/2 - (groupifs_DMSO + groupctrl_DMSO)/2,
  ctrl_PROTAC_vs_DMSO   = groupctrl_PROTAC - groupctrl_DMSO,
  stress_PROTAC_vs_DMSO = groupifs_PROTAC - groupifs_DMSO,
  interaction           = (groupifs_PROTAC - groupctrl_PROTAC) - (groupifs_DMSO - groupctrl_DMSO),
  levels = design
)

fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)  # Empirical Bayes smoothing of standard errors


# extract each contrast table 
rowData(se)$DEA.voom.total_stress_vs_ctrl <- topTable(fit2, coef = "total_stress_vs_ctrl", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.DMSO_stress_vs_ctrl <- topTable(fit2, coef = "DMSO_stress_vs_ctrl", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.PROTAC_stress_vs_ctrl <- topTable(fit2, coef = "PROTAC_stress_vs_ctrl", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.total_PROTAC_vs_DMSO <- topTable(fit2, coef = "total_PROTAC_vs_DMSO", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.ctrl_PROTAC_vs_DMSO <- topTable(fit2, coef = "ctrl_PROTAC_vs_DMSO", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.stress_PROTAC_vs_DMSO <- topTable(fit2, coef = "stress_PROTAC_vs_DMSO", n = Inf)[rownames(se),]
rowData(se)$DEA.voom.interaction <- topTable(fit2, coef = "interaction", n = Inf)

```

## Fig 14a

volcano plot of circRNA DEA results 

```{r}
res <- rowData(se)$DEA.voom.total_stress_vs_ctrl
res$host_gene <- rowData(se)[rownames(res),]$host_gene
res <- res[order(res$P.Value),]
res$sig_anno <- "non"
res$sig_anno[which(res$adj.P.Val < 0.05 & res$logFC > 0)] <- "up" # no ups
res$sig_anno[which(res$adj.P.Val < 0.05 & res$logFC< 0)] <- "down"  # no downs
tasp1_label <- head(res[which(res$host_gene == "Tasp1"), ], 1)

fig14a <- ggplot(res, aes(x = logFC, y = -log10(P.Value), label = host_gene, color = sig_anno)) + 
  geom_point()+
  geom_text_repel(
    data = tasp1_label,
    size = 8,
    color = "black",
    max.overlaps  = 20,
    box.padding   = 0.3
    ) +
  geom_point(data = tasp1_label, color = "black", inherit.aes = FALSE,aes(x = logFC, y = -log10(P.Value)))+
  scale_color_manual(name = "direction", values = c("up" = "red", "non" = "gray", "down" = "blue")) +
  labs(
    title  = "Sperm circRNA (total stress vs. ctrl)",
    x      = "logFC stress vs. ctrl",
    y      = "-log10(P.unadj)"
  ) +
  theme_bw()+
  theme(
    axis.title = element_text(size = 18),
    plot.title = element_text(hjust=0.5),
    axis.text  = element_text(size = 18),
    title  = element_text(size = 18, face = "bold"),
    legend.position = "none"
  )
fig14a

# ggsave("/mnt/gapp/work/bohan/thesis/IFS_sperm_longRNA/plot_for_thesis/volcano_circ_voom_total_stress_vs_ctrl.pdf", fig14a, width = 7, height = 7)
```

## Fig 14b 

Tasp1 circRNA

```{r}
# logcpm
tasp1 <- assay(se, "logcpm")[which(rownames(se)=="chr2:139884014|139899419"),]
df_tasp1 <- data.frame(
  sample = names(tasp1),
  stress = ifelse(substr(names(tasp1), 1, 1)=="C", "ctrl", "stress"),
  PROTAC = ifelse(substr(names(tasp1), nchar(names(tasp1)), nchar(names(tasp1)))=="D", "DMSO", "PROTAC"),
  donor = gsub("[DK]$", "", names(tasp1)),
  count = tasp1
)
df_tasp1$group <- factor(paste0(df_tasp1$stress, "_", df_tasp1$PROTAC))


fig14b <- ggplot(data = df_tasp1, aes(x = group, y = count)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.7)+
  geom_point(size = 2, alpha = 0.8) +
  geom_line(aes(group = donor), color = "grey10", alpha = 0.5, linewidth = 1.5)+
  theme_bw() +
  labs(x = "Group", y = "logcpm",  title = "Expression of Tasp1-circRNA") +
  theme(
    # axis titles
    axis.title.x  = element_blank(),
    axis.title.y  = element_text(size = 18, face = "bold"),
    title  = element_text(size = 18, face = "bold"),
    plot.title = element_text(hjust=0.5),
    # axis text
    axis.text.x   = element_text(size = 18, face = "bold", vjust = 0.1),
    axis.text.y   = element_text(size = 14),
    # axis ticks
    axis.ticks    = element_line(size = 0),
    axis.ticks.length = unit(0, "cm"),
    legend.text = element_text(size = 18)
    )
fig14b
# ggsave("./plot_for_thesis/box_tasp1_logcpm.pdf", fig14b, width = 7, height = 7)
```

