---
title: "01_arbitrary_ctrl_data"
output: html_document
author: "Bohan Zhang"
date: "2026-04-21"
---

```{r}
library(edgeR)
library(SummarizedExperiment)
library(ggplot2)
library(limma)
library(sva)
library(sechm)
library(purrr)
library(stringr)
library(tibble)
library(reshape2)
library(dplyr)
library(tidyr)
library(BiocParallel)
library(ggnewscale)
setwd("/mnt/gapp/work/bohan/thesis/SVA_exploration")
```

# data preparation

Data we use: Rebecca's CRS data

```{r}
se <- readRDS("rebecca.CRS.full.SE.rds")
# remove the previous SVs
colData(se) <- colData(se)[, !grepl("^SV", names(colData(se)))]

se <- se[, se$TimePoint %in% c("0min","24h") & se$Sex=="female"] # because we knew that 0min and 24h are not impacted by acute stress

# "Experiment": RestraintTimeline and CRS
# Here Experiment serves as a "unknown covariant" (meaning that it exists but we pretend that we don't know it)
```

Now we have 40 samples (8 RestraintTimeline, 32 CRS)

we downsample them, but the proportion of RestraintTimeline and CRS remains the same

But to get the minimum sample size as 3vs3, with 2 RestraintTimeline and 4 CRS: we use the Max sample size as 8 RestraintTimeline and 16 CRS. 

Since there are also CRS3 and CRS4, let's take 16 CRS3 for simplification

So final full dataset: 8 RestraintTimeline and 16 CRS3

```{r}
se <- se[, se$Experiment %in% c("RestraintTimeline", "CRS3")]
se$Experiment <- factor(se$Experiment)
```

# function prepare

```{r}
source("./function_for_arbitrary_ctrl_data_twoBatch_diffSampleSize.R")
source("./svacor_PL.R")
source("./proxy_function.R")
```

# test different sample sizes

We choose sample size 24, 18, 12, 6
sample size (24) = RestraintTimeline_A (4) + RestraintTimeline_B (4) + CRS_A (8) + CRS_B (8)
A vs. B = 12 vs. 12

sample size (18) = RestraintTimeline_A (3) + RestraintTimeline_B (3) + CRS_A (6) + CRS_B (6)
A vs. B = 9 vs. 9

sample size (12) = RestraintTimeline_A (2) + RestraintTimeline_B (2) + CRS_A (4) + CRS_B (4)
A vs. B = 6 vs. 6

sample size (6) = RestraintTimeline_A (1) + RestraintTimeline_B (1) + CRS_A (2) + CRS_B (8)
A vs. B = 3 vs. 3

# Run in the loop

```{r}
seedset <- c(12, 15, 48, 88, 100, 240)
sampleset <- c(24, 18, 12, 6)
bp <- MulticoreParam(length(seedset), progressbar=TRUE)

for (size in c(24, 18, 12, 6)){
    bplapply(seedset, BPPARAM=bp, FUN=function(seed){
        # subsampling and group assignment
        se_sub <- subsampling(se = se, sampleSize = size, seed = seed)

        # filter low counts
        se_sub <- filterLowCounts(se_sub)
        # calculate the logcpm
        dge <- DGEList(assay(se_sub, "counts"))
        dge <- calcNormFactors(dge)
        assay(se_sub, "logcpm") <- log1p(edgeR::cpm(dge)[rownames(se_sub),])

        # n.sv threshold
        set.seed(8)
        test1 <- svacor_PL(se_sub, ~ se_sub$group, numSVmethod = "be")
        maxSV <- sum(grepl("^SV", names(colData(test1))))
        metadata(se_sub)$"BE" <- paste0("n.sv_", maxSV)
        set.seed(24)
        test2 <- svacor_PL(se_sub, ~ se_sub$group, numSVmethod = "leek")
        minSV <- sum(grepl("^SV", names(colData(test2))))
        metadata(se_sub)$"LEEK" <- paste0("n.sv_", minSV)

        # LETS TEST ========================================================
        test_result <- list()
        for (n in 0:6){ # we use 0~6 as the range of SV numbers
            if(n < (ncol(se_sub) - 2)){  # here, for example if we have 3vs3, then if we use 4 SVs, and we also have intercept + group to estimate and the dispersion, the model is under-determined (meaning that the design doesn't have enough degrees of freedom to support so many parameters)
                test_result[[paste0("n.sv_", n)]] <- Lets_test(se = se_sub, dge = dge, n_SV = n, 
                MDS_name = paste0("./MDS_arbitrary_ctrl_data/sampleSizeTest/", size, "samples_seed_", seed))
            }
        }
        saveRDS(test_result, paste0("./test_results_twoBatch_diffSampleSize/test_results_DiffSampleSize_", size, "_seed_", seed, ".rds"))

        # P value hist
        for(i in names(test_result)){
            p <- ggplot(test_result[[i]][["DEA"]]$AvsB, aes(x = PValue)) +
                geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8, boundary = 0)+
                labs(title = paste0("pvalue histogram - ", i))
            ggsave(paste0("./pval_hist_arbitrary_ctrl_data_twoBatch_diffSampleSize/", size, "samples_seed_", seed, "_", i, ".pdf"), width = 8, height = 6)
            }
    })
}
```

# test results path

```{r}
res_path <- "./test_results_twoBatch_diffSampleSize"
files <- list.files(
  path    = res_path,
  pattern = "^test_results_DiffSampleSize_\\d+_seed_\\d+\\.rds$",
  full.names = TRUE
)
```

# Visulization parameters

```{r}
seed_colors <- c(
    "88" = "firebrick3", 
    "12" = "purple3", 
    "100" = "steelblue2", 
    "240" = "seagreen", 
    "15" = "goldenrod", 
    "48" = "deeppink"
)

seed_levels <- seedset
sample_levels <- sampleset
```

# test if p-hist is stable:

```{r}
# df_pval <- get_p_hist_stability(files, path)

# p_nsvNOTseparate <- ggplot(df_pval, aes(x = PValue, color = seed)) +
#     geom_density()+
#     facet_wrap(~sample_size) + 
#     labs(title = "Pvalue curve, n.sv not separated")
# p_nsvNOTseparate
# # ggsave("./pval_hist_arbitrary_ctrl_data_twoBatch_diffSampleSize/combined_nsvNOTseparated.pdf", p_nsvNOTseparate, width = 8, height = 6)

# p_nsvSeparate <- ggplot(df_pval, aes(x = PValue, color = seed)) +
#     geom_density()+
#     facet_wrap(~sample_size + n_sv) + 
#     labs(title = "Pvalue curve, n.sv separated")
# p_nsvSeparate
# # ggsave("./pval_hist_arbitrary_ctrl_data_twoBatch_diffSampleSize/combined_nsvSeparated.pdf", p_nsvSeparate, width = 13, height = 8)
```

# DEG number combined plot

```{r}
df_deg <- get_DEG_num(files, path)
df_deg$seed <- factor(df_deg$seed, levels = seed_levels)
df_deg$sample_size <- factor(df_deg$sample_size, levels = sample_levels)

p_deg <- ggplot(df_deg, aes(x = n_sv, y = DEG_num, fill = seed)) +
    geom_col()+
    geom_vline(aes(xintercept = sv_suggested, linetype = sv_method, color = sv_method)) + 
    scale_color_manual(name = "Suggested SV", values = c("BE" = "red", "LEEK" = "blue")) +
    scale_linetype_manual(name = "Suggested SV", values = c("BE" = "longdash", "LEEK" = "dotted")) +
    facet_grid(seed ~ sample_size, labeller = label_both) +
    labs(title = "Number of DEGs", x = "Number of SVs", y = "Number of DEGs") + 
    scale_fill_manual(values = seed_colors) + 
    theme_bw()+
    theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 12), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 14),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16))
p_deg
# ggsave("./CRS_data_combined_DEG_num.pdf", p_deg, width = 10, height = 8)
```

# correlation of SV value from n_sv.x

```{r}
# cor_all <- get_SV_value_correlation(files, path)
# cor_all$seed <- factor(cor_all$seed, levels = seed_levels)

# p_svcor <- ggplot(cor_all, aes(x = Var1, y = Var2, fill = Freq)) +
#     geom_tile() +
#     geom_text(aes(label = round(Freq, 2)), size = 3) +
#     scale_fill_gradient2(low = "cornflowerblue", mid = "white", high = "firebrick3", 
#                          midpoint = 0, limits = c(-1, 1)) +
#     facet_grid(seed ~ sample_size)+
#     labs(title = "SV Correlation Heatmap", x = "", y = "", fill = "correlation") +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_svcor
# # ggsave("./CRS_data_SV_correlation_map.pdf", p_svcor, width = 30, height = 45)
```

## curve

For each n.sv level, find the maximum correlation between the LAST SV added and all previous SVs. 

```{r}
cor_curve <- get_SV_value_correlation_curve(files, path)
cor_curve$seed <- factor(cor_curve$seed, levels = seed_levels)
cor_curve$sample_size <- factor(cor_curve$sample_size, levels = sample_levels)

p_cor_curve <- ggplot(cor_curve, aes(x = nsv, y = maxcor, color = seed, group = seed)) + 
    geom_point(size = 2) + 
    geom_line(linewidth = 1)+
    scale_color_manual(values = seed_colors) + 
    new_scale_color() + 
    geom_vline(aes(xintercept = sv_suggested, linetype = sv_method, color = sv_method)) + 
    scale_color_manual(name = "Suggested SV", values = c("BE" = "red", "LEEK" = "blue")) +
    scale_linetype_manual(name = "Suggested SV", values = c("BE" = "longdash", "LEEK" = "dotted")) +
    facet_grid(seed ~ sample_size, labeller = label_both) +
    labs(title = "SV Correlation curve", x = "Number of SVs", y = "SV Correlation Score") + 
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)), labels = scales::label_scientific(digits = 2))+
    theme_bw()+
    theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 12), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 14),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.5, "cm"))

p_cor_curve
# ggsave(paste0("./CRS_data_SV_correlation_curve.pdf"), p_cor_curve, width = 12, height = 8)

```

# DEG correlation

```{r}
N <- 1000 # for simplification

auc_all <- get_DEG_ranking_cor_auc(N, files, path)
auc_all$seed <- factor(auc_all$seed, levels = seed_levels)
auc_all$sample_size <- factor(auc_all$sample_size, levels = sample_levels)
# p_auc_deg_cor <- ggplot(auc_all, aes(x = col1, y = col2, fill = auc)) +
#     geom_tile()+
#     geom_text(aes(label = round(auc, 2)), size = 2) +
#     scale_fill_gradient2(low = "cornflowerblue", mid = "white", high = "firebrick3", 
#                          midpoint = 0.5, limits = c(0, 1)) +
#     facet_grid(seed ~ sample_size,
#                labeller = labeller(sample_size = function(x) paste("SampleSize =", x),
#                                    seed = function(x) paste("seed =", x)))+
#     labs(title = "AUC of DEG overlap curve", x = "", y = "AUC (normalized)") +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1))

# p_auc_deg_cor
# # ggsave("./CRS_data_DEGs_rank_correlation_map.pdf", p_auc_deg_cor, width = 13, height = 8)

# only show half of them in the plot (since it's symmetric)
auc_lower <- auc_all %>%
    filter(as.integer(factor(col1)) >= 
           as.integer(factor(col2)))
auc_lower$seed <- factor(auc_lower$seed, levels = seed_levels)
auc_lower$sample_size <- factor(auc_lower$sample_size, levels = sample_levels)

p_auc_deg_cor_half <- ggplot(auc_lower, aes(x = col1, y = col2, fill = auc)) +
    geom_tile()+
    geom_text(aes(label = round(auc, 2)), size = 2) +
    scale_fill_gradient2(low = "cornflowerblue", mid = "white", high = "firebrick3", 
                         midpoint = 0.5, limits = c(0, 1)) +
    facet_grid(seed ~ sample_size, labeller = label_both)+
    labs(title = "DEG Overlap Score", x = "Number of SVs", y = "Number of SVs", fill = "AUC of \noverlap curve") +
    theme_bw()+
    theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 12), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 14),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.5, "cm"))
p_auc_deg_cor_half
# ggsave("./CRS_data_DEGs_rank_correlation_map_half.pdf", p_auc_deg_cor_half, width = 10, height = 9)
```

# p value skew

```{r}

# pskew <- get_pskew(files, path)

# pskew_long <- pskew %>% pivot_longer(
#     cols = -c("sample_size", "seed"),
#     names_to = "nsv",
#     values_to = "pskew"
# )
# pskew_long$seed <- factor(pskew_long$seed)
# pskew_long$sample_size <- factor(pskew_long$sample_size)
# p_pskew <- ggplot(pskew_long, aes(x = nsv, y = seed, fill = pskew)) + 
#     geom_tile()+
#     facet_grid(~sample_size,
#                 labeller = labeller(sample_size = function(x) paste("SampleSize =", x)))+
#     scale_fill_gradient2(low = "cornflowerblue", mid = "white", high = "firebrick3") +
#     labs(fill = "mean(p)-0.5")
# p_pskew
# # ggsave("./CRS_data_p_skew_map.pdf", p_pskew, width = 13, height = 8)

```

# MDS separation

Here we check how much more the two groups are separated with the increase of the SV number

```{r}
df_pve <- get_MDS_separation(files, path)
df_pve$seed <- factor(df_pve$seed, levels = seed_levels)
df_pve$sample_size <- factor(df_pve$sample_size, levels = sample_levels)

p_pve <- ggplot(df_pve, aes(x = n_sv, y = pve, color = seed, group = seed)) + 
    geom_point(size = 2) + 
    geom_line(linewidth = 1)+
    scale_color_manual(values = seed_colors) + 
    new_scale_color() + 
    geom_vline(aes(xintercept = sv_suggested, linetype = sv_method, color = sv_method)) + 
    scale_color_manual(name = "Suggested SV", values = c("BE" = "red", "LEEK" = "blue")) +
    scale_linetype_manual(name = "Suggested SV", values = c("BE" = "longdash", "LEEK" = "dotted")) +
    facet_grid(seed ~ sample_size, labeller = label_both) + 
    labs(title = "MDS separation contribution", x = "Number of SVs", y = "Proportion of Variance Explained") +
    theme_bw()+
    theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 12), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 14),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.5, "cm"))

p_pve
# ggsave("./CRS_data_MDS_separation_contribution_curve.pdf", p_pve, width = 10, height = 8)

```

PVE > 0.5 (50%) → group separation is clear, the effect is strong
PVE 0.2–0.5 → moderate separation
PVE < 0.2 → weak or no separation

# SV contribution to correction

```{r}
df_sumAbsBs <- get_SV_contribution_to_correction(files, path)
# p_sumAbsBs <- ggplot(df_sumAbsBs, aes(x = SV_id, y = sumAbsBs, fill = SV)) +
#     geom_col() +
#     facet_grid(seed ~ sample_size, labeller = label_both) +
#     labs(title = "Contributions of each SV on the correction", y = "sumAbsBs", x = "SV") +
#     theme_bw() +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_sumAbsBs
# ggsave("./CRS_data_Contributions_of_SV_on_correction.pdf", p_sumAbsBs, width = 15, height = 12)

df_sumAbsBs_max <- df_sumAbsBs %>%
    group_by(sample_size, seed) %>%
    filter(as.integer(str_extract(n_sv, "\\d+")) == max(as.integer(str_extract(n_sv, "\\d+")))) %>%
    ungroup()
df_sumAbsBs_max$seed <- factor(df_sumAbsBs_max$seed, levels = seed_levels)
df_sumAbsBs_max$sample_size <- factor(df_sumAbsBs_max$sample_size, levels = sample_levels)

p_sumAbsBs_max <- ggplot(df_sumAbsBs_max, aes(x = SV, y = sumAbsBs, fill = seed)) +
    geom_col() +
    facet_grid(seed ~ sample_size, labeller = label_both) +
    labs(title = "Contributions of each SV on the correction", y = "sumAbsBs", x = "SV") +
    scale_fill_manual(values = seed_colors) + 
    theme_bw()+
    theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 12), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 14),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.5, "cm"))

p_sumAbsBs_max
# ggsave("./CRS_data_Contributions_of_SV_on_correction_maxSVnum.pdf", p_sumAbsBs_max, width = 10, height = 8)
```

# SV contribution to logFC of the DEGs

```{r}
df_SV_contribute <- get_SV_contribution_to_logFC_DEG(files, path)
df_SV_contribute$seed <- factor(df_SV_contribute$seed, levels = seed_levels)
df_SV_contribute$sample_size <- factor(df_SV_contribute$sample_size, levels = sample_levels)

p_SV_contribute <- ggplot(df_SV_contribute, aes(x = n_sv, y = SV_contribute, color = seed, group = seed)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_manual(values = seed_colors) + 
    new_scale_color() + 
    geom_vline(aes(xintercept = sv_suggested, linetype = sv_method, color = sv_method)) + 
    scale_color_manual(name = "Suggested SV", values = c("BE" = "red", "LEEK" = "blue")) +
    scale_linetype_manual(name = "Suggested SV", values = c("BE" = "longdash", "LEEK" = "dotted")) +
    facet_grid(seed ~ sample_size, labeller = label_both) +
    labs(title = "SV contribution to logFC", x = "Number of SVs", y = "DEGs, Max mean(logFC_SV - logFC_AvsB)", color = "seed") +
    theme_bw()+
    theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 12), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 14),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.5, "cm"))
p_SV_contribute
# ggsave("./CRS_data_SV_contribute_logFC.pdf", p_SV_contribute, width = 10, height = 8)
```

The empty value means that in this case there is no DEGs detected

# overlap between SV sig rank and AvsB sig rank

The concept: when it increases, there is a problem because the significance are more from SV rather than the real effect

```{r}
N <- 1000
sig_rank_cor <- get_significance_ranking_cor_auc(N, files, path)
sig_rank_cor$seed <- factor(sig_rank_cor$seed, levels = seed_levels)
sig_rank_cor$sample_size <- factor(sig_rank_cor$sample_size, levels = sample_levels)

p_sig_rank_cor <- ggplot(sig_rank_cor, aes(x = n_sv, y = auc, group = seed, color = seed)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    facet_grid(seed ~ sample_size, labeller = label_both) +
    scale_color_manual(values = seed_colors) + 
    new_scale_color() + 
    geom_vline(data = ~ filter(.x, !is.na(sv_suggested)), aes(xintercept = sv_suggested, linetype = sv_method, color = sv_method)) + 
    scale_color_manual(name = "Suggested SV", values = c("BE" = "red", "LEEK" = "blue")) +
    scale_linetype_manual(name = "Suggested SV", values = c("BE" = "longdash", "LEEK" = "dotted")) +
    labs(title = "Significance ranking correlation\nbetween SV effect and true DEGs", x = "Number of SVs", y = "auc", color = "seed") +
    theme_bw()+
    theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 12), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 14),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.3, "cm"))
p_sig_rank_cor
# ggsave("./CRS_data_noDEG_sig_rank_cor.pdf", p_sig_rank_cor, width = 10, height = 8)
```

# some MDS example:

```{r}
test_samplesize_24_seed_48 <- readRDS("/mnt/gapp/work/bohan/thesis/SVA_exploration/test_results_twoBatch_diffSampleSize/test_results_DiffSampleSize_24_seed_48.rds")
```

## logcpm

```{r}
se_nsv0 <- test_samplesize_24_seed_48$n.sv_0$SE
groupcol <- se_nsv0$group
groupcol <- c(A = "orange2", B = "navy")[groupcol]
pdf("24samples_seed_48_logcpm_exampleUse_CRS.pdf", width = 8, height = 6, bg = "white")
limma::plotMDS(assay(se_nsv0, "logcpm"), labels = paste0(se_nsv0$group, "_", se_nsv0$Experiment), col = groupcol, main = "MDS: 24samples_seed_48_logcpm")
dev.off()

```

## n.sv = 1

```{r}
se_nsv1 <- test_samplesize_24_seed_48$n.sv_1$SE
groupcol <- se_nsv1$group
groupcol <- c(A = "orange2", B = "navy")[groupcol]
pdf("24samples_seed_48_nsv.1_exampleUse_CRS.pdf", width = 8, height = 6, bg = "white")
limma::plotMDS(assay(se_nsv1, "corrected"), labels = paste0(se_nsv1$group, "_", se_nsv1$Experiment), col = groupcol, main = "MDS: 24samples_seed_48_nsv.1_corrected")
dev.off()

```

## n.sv = 2

```{r}
se_nsv2 <- test_samplesize_24_seed_48$n.sv_2$SE
groupcol <- se_nsv2$group
groupcol <- c(A = "orange2", B = "navy")[groupcol]
pdf("24samples_seed_48_nsv.2_exampleUse_CRS.pdf", width = 8, height = 6, bg = "white")
limma::plotMDS(assay(se_nsv2, "corrected"), labels = paste0(se_nsv2$group, "_", se_nsv2$Experiment), col = groupcol, main = "MDS: 24samples_seed_48_nsv.2_corrected")
dev.off()

```

## n.sv = 6

```{r}
se_nsv6 <- test_samplesize_24_seed_48$n.sv_6$SE
groupcol <- se_nsv6$group
groupcol <- c(A = "orange2", B = "navy")[groupcol]
pdf("24samples_seed_48_nsv.6_exampleUse_CRS.pdf", width = 8, height = 6, bg = "white")
limma::plotMDS(assay(se_nsv6, "corrected"), labels = paste0(se_nsv6$group, "_", se_nsv6$Experiment), col = groupcol, main = "MDS: 24samples_seed_48_nsv.6_corrected")
dev.off()

```

# 6 samples

```{r}
test_samplesize_6_seed_48 <- readRDS("/mnt/gapp/work/bohan/thesis/SVA_exploration/test_results_twoBatch_diffSampleSize/test_results_DiffSampleSize_6_seed_48.rds")
```

## logcpm

```{r}
se_nsv0 <- test_samplesize_6_seed_48$n.sv_0$SE
groupcol <- se_nsv0$group
groupcol <- c(A = "orange2", B = "navy")[groupcol]
pdf("6samples_seed_48_logcpm_exampleUse_CRS.pdf", width = 8, height = 6, bg = "white")
limma::plotMDS(assay(se_nsv0, "logcpm"), labels = paste0(se_nsv0$group, "_", se_nsv0$Experiment), col = groupcol, main = "MDS: 6samples_seed_48_logcpm")
dev.off()

```

## n.sv = 1

```{r}
se_nsv1 <- test_samplesize_6_seed_48$n.sv_1$SE
groupcol <- se_nsv1$group
groupcol <- c(A = "orange2", B = "navy")[groupcol]
pdf("6samples_seed_48_nsv.1_exampleUse_CRS.pdf", width = 8, height = 6, bg = "white")
limma::plotMDS(assay(se_nsv1, "corrected"), labels = paste0(se_nsv1$group, "_", se_nsv1$Experiment), col = groupcol, main = "MDS: 6samples_seed_48_nsv.1_corrected")
dev.off()

```

## n.sv = 2

```{r}
se_nsv2 <- test_samplesize_6_seed_48$n.sv_2$SE
groupcol <- se_nsv2$group
groupcol <- c(A = "orange2", B = "navy")[groupcol]
pdf("6samples_seed_48_nsv.2_exampleUse_CRS.pdf", width = 8, height = 6, bg = "white")
limma::plotMDS(assay(se_nsv2, "corrected"), labels = paste0(se_nsv2$group, "_", se_nsv2$Experiment), col = groupcol, main = "MDS: 6samples_seed_48_nsv.2_corrected")
dev.off()

```

# combine the proxy curves

```{r}
# ── 1. DEG count ─────────────────────────────────────────────────────────────
df_deg_long <- df_deg %>%
    dplyr::select(sample_size, seed, n_sv, DEG_num, sv_method, sv_suggested) %>%
    rename(value = DEG_num) %>%
    mutate(proxy = "DEG Count")

# ── 2. AUC: nsv_k vs nsv_{k-1} (neighbor stability) ────────────────────────
# Re-derive from auc_all: keep only consecutive-neighbor pairs
auc_neighbor <- auc_all %>%
    mutate(
        k1 = as.integer(str_extract(col1, "\\d+")),
        k2 = as.integer(str_extract(col2, "\\d+"))
    ) %>%
    filter(k2 == k1 - 1) %>%                          # col1 = nsv_k, col2 = nsv_{k-1}
    mutate(n_sv = paste0("n.sv_", k1)) %>%             # assign to the higher nsv level
    dplyr::select(sample_size, seed, n_sv, auc) %>%
    rename(value = auc) %>%
    mutate(proxy = "DEG Overlap\n(vs prev SV)")

# ── 3. MDS separation (PVE) ──────────────────────────────────────────────────
df_pve_long <- df_pve %>%
    dplyr::select(sample_size, seed, n_sv, pve, sv_method, sv_suggested) %>%
    rename(value = pve) %>%
    mutate(proxy = "MDS Separation\n(PVE)")

# ── 5. SV contribution ───────────────────────────────────────────────────────────────
df_sumAbsBs_long <- df_sumAbsBs_max %>%
    dplyr::mutate(
        n_sv  = paste0("n.sv_", str_extract(SV, "\\d+"))
    ) %>%
    dplyr::rename(value = sumAbsBs) %>%
    dplyr::select(sample_size, seed, n_sv, value) %>%
    dplyr::mutate(
        proxy        = "SV Contribution\n(sumAbsBs)",
        sv_method    = NA_character_,
        sv_suggested = NA_character_
    )

# ── 6. Bind proxies that carry vline info, add empty vline cols to auc ───────
vline_proxies <- bind_rows(df_deg_long, df_pve_long)
# Extract one vline table (same across proxies — take from df_deg_long)
vlines <- df_deg_long %>%
    dplyr::select(sample_size, seed, sv_method, sv_suggested) %>%
    distinct()

auc_neighbor <- auc_neighbor %>%
    mutate(sv_method = NA_character_, sv_suggested = NA_character_)

all_proxies_raw <- bind_rows(vline_proxies, auc_neighbor, df_sumAbsBs_long)

# ── 7. Min-max normalise within each panel × proxy ───────────────────────────
all_proxies <- all_proxies_raw %>%
    group_by(sample_size, seed, proxy) %>%
    mutate(
        vmax  = max(value, na.rm = TRUE),
        value_norm = value / (vmax+0.01)   # here add 0.01 to prevent vmax=0
    ) %>%
    ungroup() %>%
    mutate(
        sample_size = factor(sample_size, levels = sample_levels),
        seed        = factor(seed,        levels = seed_levels),
        proxy       = factor(proxy, levels = c("DEG Count",
                                               "DEG Overlap\n(vs prev SV)",
                                               "MDS Separation\n(PVE)",
                                               "SV Contribution\n(sumAbsBs)"))
    )

# ── 7. Proxy colour palette ───────────────────────────────────────────────────
proxy_colors <- c(
    "DEG Count"              = "#cf191c",
    "DEG Overlap\n(vs prev SV)" = "#f9b52d",
    "MDS Separation\n(PVE)" = "#19a415",
    "SV Contribution\n(sumAbsBs)" = "#992aba"
)

# ── 8. Combined plot ──────────────────────────────────────────────────────────
p_combined <- ggplot(all_proxies, aes(x = n_sv, y = value_norm, color = proxy, group = proxy)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.6) +
    scale_color_manual(
        name   = "Proxy",
        values = proxy_colors
    ) +
    # BE / LEEK suggested-SV vlines
    new_scale_color()+
    geom_vline(
        data = vlines %>%
            mutate(sample_size = factor(sample_size, levels = sample_levels),
                   seed        = factor(seed,        levels = seed_levels)) %>%
            filter(!is.na(sv_suggested)),
        aes(xintercept = sv_suggested,
            linetype   = sv_method,
            color      = sv_method),          # reuse color aes — add to scale below
        inherit.aes = FALSE
    ) +
    scale_color_manual(
        name   = "Suggested SV",
        values = c("BE" = "red", "LEEK" = "blue")
    ) +
    scale_linetype_manual(
        name   = "Suggested SV",
        values = c("BE" = "longdash", "LEEK" = "dotted")
    ) +
    # keep proxy lines solid; vlines get their own linetypes via the data
    facet_grid(seed ~ sample_size, labeller = label_both) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
    labs(
        title = "CRS with one confounder - Combined proxy curves (max normalized)",
        x     = "Number of SVs",
        y     = "Normalised value"
    ) +
    theme_bw() +
    theme(
        axis.text.x    = element_text(size = 11, angle = 45, hjust = 1),
        axis.text.y    = element_text(size = 11),
        strip.text.x   = element_text(size = 13),
        strip.text.y   = element_text(size = 13),
        axis.title.x   = element_text(size = 13),
        axis.title.y   = element_text(size = 13),
        plot.title     = element_text(size = 15),
        legend.title = element_text(face = "bold"),
        legend.position = "bottom",
        panel.spacing.y = unit(0.4, "cm")
    ) +
    guides(
        color    = guide_legend(nrow = 1),
        linetype = guide_legend(nrow = 1)
    )

p_combined
ggsave("./CRS_data_combine.pdf", p_combined, width = 10, height = 8)
```