---
title: "02_spike_in_data_real_data_only_Sex_confounder"
output: html_document
author: "Bohan Zhang"
date: "2026-04-27"
---

```{r}
library(edgeR)
library(SummarizedExperiment)
library(ggplot2)
library(limma)
library(sva)
library(sechm)
library(purrr)
library(stringr)
library(tibble)
library(reshape2)
library(dplyr)
library(tidyr)
library(iPSCpoweR)
library(BiocParallel)
library(ggnewscale)
setwd("/mnt/gapp/work/bohan/thesis/SVA_exploration")
```

# function prepare

```{r}
source("function_for_spike_in_data_two_confounder.R")
source("svacor_PL.R")
source("proxy_function.R")
```

# data preparation

```{r}
data("GSE79636")
se <- SummarizedExperiment(list(counts=GSE79636$dat), colData=GSE79636$annotation)
se <- se[,se$race=="White"]  # keep Sex and insulin resistance as two confounders
se <- se[rowSums(assay(se)>10)>1,]
table(se$Sex, se$state)
#          Insulin resistant Insulin sensitive
#   Female                21                20
#   Male                  25                 9
se$confounders <- paste0(se$Sex, "_", ifelse(se$state == "Insulin resistant", "InsulRes", "InsulSens"))
table(se$confounders)
```

# run in the loop

```{r}
# set multiple cores to make it faster
seedset <- c(8, 12, 15, 32, 36, 42, 58, 88, 188, 259, 345, 356, 564, 757, 837, 999)  # 16 seeds
sampleset <- c(24, 18, 12, 6)

bp <- MulticoreParam(length(seedset), progressbar=TRUE)

for (size in c(24, 18, 12, 6)){
    bplapply(seedset, BPPARAM=bp, FUN=function(seed){
        # subsampling and group assignment
        se_sub <- subsampling(se = se, sampleSize = size, seed = seed)

        # filter low counts
        se_sub <- filterLowCounts(se_sub)
        # calculate the logcpm
        dge <- DGEList(assay(se_sub, "counts"))
        dge <- calcNormFactors(dge)
        assay(se_sub, "logcpm") <- log1p(edgeR::cpm(dge)[rownames(se_sub),])

        # n.sv threshold
        set.seed(8)
        test1 <- svacor_PL(se_sub, ~ se_sub$group, numSVmethod = "be")
        maxSV <- sum(grepl("^SV", names(colData(test1))))
        metadata(se_sub)$"BE" <- paste0("n.sv_", maxSV)

        test2 <- tryCatch({  # here we use this because the SVA sometimes recommend too much SVs, and it gives error when sample size is too low
            set.seed(24)
            test2 <- svacor_PL(se_sub, ~ se_sub$group, numSVmethod = "leek")
            minSV <- sum(grepl("^SV", names(colData(test2))))
            metadata(se_sub)$"LEEK" <- paste0("n.sv_", minSV)
            test2},
            error = function(e) {
                cat("for sample size", size, ", svacor failed: ", e$message, " — skipping\n", 
                    file = "OUTPUT_spike_in_data_simu.txt", 
                    append=TRUE)
                NULL  # returns NULL instead of crashing
            })
        if (is.null(test2)) {metadata(se_simu)$LEEK <- NA}
            
        # LETS TEST ========================================================
        test_result <- list()
        for (n in 0:6){  # It doesn't make sense to use too much SVs so lets restrict the maximum to 6 SVs
            if(n < (ncol(se_sub) - 2)){  # here, for example if we have 3vs3, then if we use 4 SVs, and we also have intercept + group to estimate and the dispersion, the model is under-determined (meaning that the design doesn't have enough degrees of freedom to support so many parameters)
            test_result[[paste0("n.sv_", n)]] <- Lets_test(se = se_sub, dge = dge, n_SV = n,
            MDS_name = paste0("./MDS_spike_in_data_two_confounder/", size, "samples_seed_", seed))
            }
        }
        saveRDS(test_result, paste0("./test_results_spike_in_data_two_confounder/test_results_DiffSampleSize_", size, "_seed_", seed, ".rds"))
        
        # P value hist
        for(i in names(test_result)){
            p <- ggplot(test_result[[i]][["DEA"]]$AvsB, aes(x = PValue)) +
                geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8, boundary = 0)+
                labs(title = paste0("pvalue histogram - ", i))
            ggsave(paste0("./pval_hist_spike_in_data_two_confounder/", size, "samples_seed_", seed, "_", i, ".pdf"), width = 8, height = 6)
        }
    })
}

```

# test results path

```{r}
res_path <- paste0("./test_results_spike_in_data_two_confounder")
files <- list.files(
    path    = res_path,
    pattern = "^test_results_DiffSampleSize_\\d+_seed_\\d+\\.rds$",
    full.names = TRUE
)
```

# Visulization parameters

```{r}
seed_colors <- setNames(
    c("steelblue", "tomato", "seagreen", "darkorchid", "goldenrod", "deeppink",
    "dodgerblue", "orangered", "mediumseagreen", "mediumpurple", "chocolate",
    "cyan4", "firebrick", "royalblue", "darkolivegreen", "hotpink"),
    seedset
)
seed_levels <- seedset
sample_levels <- sampleset

```

# DEG number combined plot

```{r}
df_deg <- get_DEG_num(files, path)
df_deg$seed <- factor(df_deg$seed, levels = seed_levels)
df_deg$sample_size <- factor(df_deg$sample_size, levels = sample_levels)

p_deg <- ggplot(df_deg, aes(x = n_sv, y = DEG_num, fill = seed)) +
    geom_col()+
    geom_vline(aes(xintercept = sv_suggested, linetype = sv_method, color = sv_method)) + 
    scale_color_manual(name = "Suggested SV", values = c("BE" = "red", "LEEK" = "blue")) +
    scale_linetype_manual(name = "Suggested SV", values = c("BE" = "longdash", "LEEK" = "dotted")) +
    facet_grid(seed ~ sample_size, labeller = label_both) +
    labs(title = "Number of DEGs", x = "Number of SVs", y = "Number of DEGs") + 
    scale_fill_manual(values = seed_colors) + 
    theme_bw()+
    theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 10), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.2, "cm"))
p_deg

# ggsave("./spike_in_two_confounder_combined_DEG_num.pdf", p_deg, width = 10, height = 16)

```

# DEG correlation

```{r}
N <- 1000 

auc_all <- get_DEG_ranking_cor_auc(N, files, path)
auc_all$seed <- factor(auc_all$seed, levels = seed_levels)
auc_all$sample_size <- factor(auc_all$sample_size, levels = sample_levels)
# p_auc_deg_cor <- ggplot(auc_all, aes(x = col1, y = col2, fill = auc)) +
#     geom_tile()+
#     geom_text(aes(label = round(auc, 2)), size = 2) +
#     scale_fill_gradient2(low = "cornflowerblue", mid = "white", high = "firebrick3", midpoint = 0.5, limits = c(0, 1)) +
#     facet_grid(seed ~ sample_size,
#                labeller = labeller(sample_size = function(x) paste("SampleSize =", x),
#                                seed = function(x) paste("seed =", x)))+
#     labs(title = "AUC of DEG overlap curve", x = "", y = "AUC (normalized)") +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1))

# p_auc_deg_cor
# ggsave("./spike_in_two_confounder_DEGs_rank_correlation_map.pdf", p_auc_deg_cor, width = 13, height = 13)

# only show half of them in the plot (since it's symmetric)
auc_lower <- auc_all %>%
    filter(as.integer(factor(col1)) >= as.integer(factor(col2)))
auc_lower$seed <- factor(auc_lower$seed, levels = seed_levels)
auc_lower$sample_size <- factor(auc_lower$sample_size, levels = sample_levels)

p_auc_deg_cor_half <- ggplot(auc_lower, aes(x = col1, y = col2, fill = auc)) +
    geom_tile()+
    geom_text(aes(label = round(auc, 2)), size = 2) +
    scale_fill_gradient2(low = "cornflowerblue", mid = "white", high = "firebrick3", 
                         midpoint = 0.5, limits = c(0, 1)) +
    facet_grid(seed ~ sample_size, labeller = label_both)+
    labs(title = "DEG Overlap Score", x = "Number of SVs", y = "Number of SVs", fill = "AUC of \noverlap curve") +
    theme_bw()+
    theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 10), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.2, "cm"))
p_auc_deg_cor_half

# ggsave("./spike_in_two_confounder_DEGs_rank_correlation_map_half.pdf", p_auc_deg_cor_half, width = 10, height = 16)

```

# p value skew

```{r}
# pskew <- get_pskew(files, path)

# pskew_long <- pskew %>% pivot_longer(
#     cols = -c("sample_size", "seed"),
#     names_to = "nsv",
#     values_to = "pskew"
# )
# pskew_long$seed <- factor(pskew_long$seed)
# pskew_long$sample_size <- factor(pskew_long$sample_size, levels = c(24,18,12,6))
# p_pskew <- ggplot(pskew_long, aes(x = nsv, y = seed, fill = pskew)) + 
#     geom_tile()+
#     facet_grid(~sample_size,
#             labeller = labeller(sample_size = function(x) paste("SampleSize =", x)))+
#     scale_fill_gradient2(low = "cornflowerblue", mid = "white", high = "firebrick3") +
#     labs(fill = "mean(p)-0.5")
# p_pskew
# # ggsave(paste0("./spike_in_two_confounder_p_skew_map.pdf"), p_pskew, width = 13, height = 8)

```

# correlation of SV value from n_sv.x

the correlation between each SVs in one SVA correction

```{r}
# cor_all <- get_SV_value_correlation(files, path)

# p_svcor <- ggplot(cor_all, aes(x = Var1, y = Var2, fill = Freq)) +
#     geom_tile() +
#     geom_text(aes(label = round(Freq, 2)), size = 3) +
#     scale_fill_gradient2(low = "cornflowerblue", mid = "white", high = "firebrick3", 
#                      midpoint = 0, limits = c(-1, 1)) +
#     facet_grid(seed ~ sample_size)+
#     labs(title = "SV Correlation Heatmap", x = "", y = "", fill = "correlation") +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_svcor
# # ggsave(paste0("./spike_in_two_confounder_SV_correlation_map.pdf"), p_svcor, width = 30, height = 45)

```

## curve

For each n.sv level, find the maximum correlation between the LAST SV added and all previous SVs. 

```{r}
cor_curve <- get_SV_value_correlation_curve(files, path)
cor_curve$seed <- factor(cor_curve$seed, levels = seed_levels)
cor_curve$sample_size <- factor(cor_curve$sample_size, levels = sample_levels)

p_cor_curve <- ggplot(cor_curve, aes(x = nsv, y = maxcor, color = seed, group = seed)) + 
    geom_point(size = 2) + 
    geom_line(linewidth = 1)+
    scale_color_manual(values = seed_colors) + 
    new_scale_color() + 
    geom_vline(aes(xintercept = sv_suggested, linetype = sv_method, color = sv_method)) + 
    scale_color_manual(name = "Suggested SV", values = c("BE" = "red", "LEEK" = "blue")) +
    scale_linetype_manual(name = "Suggested SV", values = c("BE" = "longdash", "LEEK" = "dotted")) +
    facet_grid(seed ~ sample_size, labeller = label_both) +
    labs(title = "SV Correlation curve", x = "Number of SVs", y = "SV Correlation Score") + 
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)), labels = scales::label_scientific(digits = 2))+
    theme_bw()+
    theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 10), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.4, "cm"))

p_cor_curve

# ggsave(paste0("./spike_in_two_confounder_SV_correlation_curve.pdf"), p_cor_curve, width = 10, height = 16)

```

# MDS separation

Here we check how much more the two groups are separated with the increase of the SV number

```{r}
df_pve <- get_MDS_separation(files, path)

df_pve$sample_size <- factor(df_pve$sample_size, levels = c(24, 18, 12 ,6))
df_pve$seed <- factor(df_pve$seed, levels = seed_levels)

p_pve <- ggplot(df_pve, aes(x = n_sv, y = pve, color = seed, group = seed)) + 
    geom_point(size = 2) + 
    geom_line(linewidth = 1)+
    scale_color_manual(values = seed_colors) + 
    new_scale_color() + 
    geom_vline(aes(xintercept = sv_suggested, linetype = sv_method, color = sv_method)) + 
    scale_color_manual(name = "Suggested SV", values = c("BE" = "red", "LEEK" = "blue")) +
    scale_linetype_manual(name = "Suggested SV", values = c("BE" = "longdash", "LEEK" = "dotted")) +
    facet_grid(seed ~ sample_size, labeller = label_both) + 
    labs(title = "MDS separation contribution", x = "Number of SVs", y = "Proportion of Variance Explained") +
    theme_bw()+
    theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 10), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.3, "cm"))

p_pve
# ggsave("./spike_in_two_confounder_MDS_separation_contribution_curve.pdf", p_pve, width = 10, height = 16)

```

# SV contribution to correction

```{r}
df_sumAbsBs <- get_SV_contribution_to_correction(files, path)

# p_sumAbsBs <- ggplot(df_sumAbsBs, aes(x = SV_id, y = sumAbsBs, fill = SV)) +
#     geom_col() +
#     facet_grid(seed ~ sample_size, labeller = label_both) +
#     labs(title = "Contributions of each SV on the correction", y = "sumAbsBs", x = "SV") +
#     theme_bw() +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_sumAbsBs
# # ggsave("./spike_in_two_confounder_Contributions_of_SV_on_correction.pdf", p_sumAbsBs, width = 15, height = 12)

df_sumAbsBs_max <- df_sumAbsBs %>%
    group_by(sample_size, seed) %>%
    filter(as.integer(str_extract(n_sv, "\\d+")) == max(as.integer(str_extract(n_sv, "\\d+")))) %>%
    ungroup()
df_sumAbsBs_max$seed <- factor(df_sumAbsBs_max$seed, levels = seed_levels)
df_sumAbsBs_max$sample_size <- factor(df_sumAbsBs_max$sample_size, levels = sample_levels)

p_sumAbsBs_max <- ggplot(df_sumAbsBs_max, aes(x = SV, y = sumAbsBs, fill = seed)) +
    geom_col() +
    facet_grid(seed ~ sample_size, labeller = label_both) +
    labs(title = "Contributions of each SV on the correction", y = "sumAbsBs", x = "SV") +
    scale_fill_manual(values = seed_colors) + 
    theme_bw()+
    theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 10), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.3, "cm"))

p_sumAbsBs_max
# ggsave("./spike_in_two_confounder_Contributions_of_SV_on_correction_maxSVnum.pdf", p_sumAbsBs_max, width = 10, height = 16)
```

# SV contribution to logFC of the DEGs

```{r}
df_SV_contribute <- get_SV_contribution_to_logFC_DEG(files, path)
df_SV_contribute$seed <- factor(df_SV_contribute$seed, levels = seed_levels)
df_SV_contribute$sample_size <- factor(df_SV_contribute$sample_size, levels = sample_levels)

p_SV_contribute <- ggplot(df_SV_contribute, aes(x = n_sv, y = SV_contribute, color = seed, group = seed)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_manual(values = seed_colors) + 
    new_scale_color() + 
    geom_vline(aes(xintercept = sv_suggested, linetype = sv_method, color = sv_method)) + 
    scale_color_manual(name = "Suggested SV", values = c("BE" = "red", "LEEK" = "blue")) +
    scale_linetype_manual(name = "Suggested SV", values = c("BE" = "longdash", "LEEK" = "dotted")) +
    facet_grid(seed ~ sample_size, labeller = label_both) +
    labs(title = "SV contribution to logFC", x = "Number of SVs", y = "DEGs, Max mean(logFC_SV - logFC_AvsB)", color = "seed") +
    theme_bw()+
    theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
          axis.text.y = element_text(size = 10), 
          strip.text.x = element_text(size = 14), 
          strip.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title = element_text(size = 16),
          panel.spacing.y = unit(0.3, "cm"))
p_SV_contribute
# ggsave("./spike_in_two_confounder_SV_contribute_logFC.pdf", p_SV_contribute, width = 10, height = 16)
```

# combine the proxy curves

```{r}
# ── 1. DEG count ─────────────────────────────────────────────────────────────
df_deg_long <- df_deg %>%
    dplyr::select(sample_size, seed, n_sv, DEG_num, sv_method, sv_suggested) %>%
    rename(value = DEG_num) %>%
    mutate(proxy = "DEG Count")

# ── 2. AUC: nsv_k vs nsv_{k-1} (neighbor stability) ────────────────────────
# Re-derive from auc_all: keep only consecutive-neighbor pairs
auc_neighbor <- auc_all %>%
    mutate(
        k1 = as.integer(str_extract(col1, "\\d+")),
        k2 = as.integer(str_extract(col2, "\\d+"))
    ) %>%
    filter(k2 == k1 - 1) %>%                          # col1 = nsv_k, col2 = nsv_{k-1}
    mutate(n_sv = paste0("n.sv_", k1)) %>%             # assign to the higher nsv level
    dplyr::select(sample_size, seed, n_sv, auc) %>%
    rename(value = auc) %>%
    mutate(proxy = "DEG Overlap\n(vs prev SV)")

# ── 3. MDS separation (PVE) ──────────────────────────────────────────────────
df_pve_long <- df_pve %>%
    dplyr::select(sample_size, seed, n_sv, pve, sv_method, sv_suggested) %>%
    rename(value = pve) %>%
    mutate(proxy = "MDS Separation\n(PVE)")

# ── 5. SV contribution ───────────────────────────────────────────────────────────────
df_sumAbsBs_long <- df_sumAbsBs_max %>%
    dplyr::mutate(
        n_sv  = paste0("n.sv_", str_extract(SV, "\\d+"))
    ) %>%
    dplyr::rename(value = sumAbsBs) %>%
    dplyr::select(sample_size, seed, n_sv, value) %>%
    dplyr::mutate(
        proxy        = "SV Contribution\n(sumAbsBs)",
        sv_method    = NA_character_,
        sv_suggested = NA_character_
    )

# ── 6. Bind proxies that carry vline info, add empty vline cols to auc ───────
vline_proxies <- bind_rows(df_deg_long, df_pve_long)
# Extract one vline table (same across proxies — take from df_deg_long)
vlines <- df_deg_long %>%
    dplyr::select(sample_size, seed, sv_method, sv_suggested) %>%
    distinct()

auc_neighbor <- auc_neighbor %>%
    mutate(sv_method = NA_character_, sv_suggested = NA_character_)

all_proxies_raw <- bind_rows(vline_proxies, auc_neighbor, df_sumAbsBs_long)

# ── 7. Min-max normalise within each panel × proxy ───────────────────────────
all_proxies <- all_proxies_raw %>%
    group_by(sample_size, seed, proxy) %>%
    mutate(
        vmax  = max(value, na.rm = TRUE),
        value_norm = value / (vmax+0.01)   # here add 0.01 to prevent vmax=0
    ) %>%
    ungroup() %>%
    mutate(
        sample_size = factor(sample_size, levels = sample_levels),
        seed        = factor(seed,        levels = seed_levels),
        proxy       = factor(proxy, levels = c("DEG Count",
                                               "DEG Overlap\n(vs prev SV)",
                                               "MDS Separation\n(PVE)",
                                               "SV Contribution\n(sumAbsBs)"))
    )

# ── 7. Proxy colour palette ───────────────────────────────────────────────────
proxy_colors <- c(
    "DEG Count"              = "#cf191c",
    "DEG Overlap\n(vs prev SV)" = "#f9b52d",
    "MDS Separation\n(PVE)" = "#19a415",
    "SV Contribution\n(sumAbsBs)" = "#992aba"
)

# ── 8. Combined plot ──────────────────────────────────────────────────────────
p_combined <- ggplot(all_proxies, aes(x = n_sv, y = value_norm, color = proxy, group = proxy)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.6) +
    scale_color_manual(
        name   = "Proxy",
        values = proxy_colors
    ) +
    # BE / LEEK suggested-SV vlines
    new_scale_color()+
    geom_vline(
        data = vlines %>%
            mutate(sample_size = factor(sample_size, levels = sample_levels),
                   seed        = factor(seed,        levels = seed_levels)) %>%
            filter(!is.na(sv_suggested)),
        aes(xintercept = sv_suggested,
            linetype   = sv_method,
            color      = sv_method),          # reuse color aes — add to scale below
        inherit.aes = FALSE
    ) +
    scale_color_manual(
        name   = "Suggested SV",
        values = c("BE" = "red", "LEEK" = "blue")
    ) +
    scale_linetype_manual(
        name   = "Suggested SV",
        values = c("BE" = "longdash", "LEEK" = "dotted")
    ) +
    # keep proxy lines solid; vlines get their own linetypes via the data
    facet_grid(seed ~ sample_size, labeller = label_both) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
    labs(
        title = "iPSC with two confounders - Combined proxy curves (max normalized)",
        x     = "Number of SVs",
        y     = "Normalised value"
    ) +
    theme_bw() +
    theme(
        axis.text.x    = element_text(size = 11, angle = 45, hjust = 1),
        axis.text.y    = element_text(size = 11),
        strip.text.x   = element_text(size = 13),
        strip.text.y   = element_text(size = 13),
        axis.title.x   = element_text(size = 13),
        axis.title.y   = element_text(size = 13),
        plot.title     = element_text(size = 15),
        legend.title = element_text(face = "bold"),
        legend.position = "bottom",
        panel.spacing.y = unit(0.4, "cm")
    ) +
    guides(
        color    = guide_legend(nrow = 1),
        linetype = guide_legend(nrow = 1)
    )

p_combined
ggsave("./spike_in_two_confounder_combine.pdf", p_combined, width = 10, height = 18)
```

